Exploratory_Analysis_Josh_Tobey

### 2025-12-09 Exploratory_Analysis_Josh_Tobey.Rmd
Updated version of GAM so far for each predictor.

coagual

>>>>

```{r}
ggplot(exploratory_df) +
  geom_point(aes(x = mean_pH,
                 y = mean_Temp))
```
```{r}
RWAR_dosing_lag_incdec <- read_csv("RWAR_dosing_lag_incdec.csv")
head(RWAR_dosing_lag_incdec)

```
```{r}
RWAR_dosing_lag_incdec <- RWAR_dosing_lag_incdec %>%
  select(-c(FeCl_change, Polymer_change, Timestamp))
head(RWAR_dosing_lag_incdec)
```

```{r}
library(mgcv)

lm_spec <- linear_reg()

gam_formula <- Ferric.chloride..mg.L ~ s(Cationic.Polimer..mg.L) + s(mean_Temp) + s(mean_pH) + s(mean_Conductivity) + s(mean_Turbidity) + s(Temp_lag1) + s(Temp_lag2) + s(pH_lag1) + s(pH_lag2) + s(Cond_lag1) + s(Cond_lag2) + s(Turb_lag1) + s(Turb_lag2)

gam_fit <- gam(gam_formula, data = RWAR_dosing_lag_incdec)
gam_rec <- recipe(Ferric.chloride..mg.L ~ ., data = RWAR_dosing_lag_incdec) %>%
  step_ns(Cationic.Polimer..mg.L, deg_free = 10) %>%
  step_ns(mean_Temp, deg_free = 10) %>%
  step_ns(mean_pH, deg_free = 10) %>%
  step_ns(mean_Conductivity, deg_free = 10) %>%
  step_ns(mean_Turbidity, deg_free = 10) %>%
  step_ns(Temp_lag1, deg_free = 10) %>%
  step_ns(Temp_lag2, deg_free = 10) %>%
  step_ns(pH_lag1, deg_free = 10) %>%
  step_ns(pH_lag2, deg_free = 10) %>%
  step_ns(Cond_lag1, deg_free = 10) %>%
  step_ns(Cond_lag2, deg_free = 10) %>%
  step_ns(Turb_lag1, deg_free = 10) %>%
  step_ns(Turb_lag2, deg_free = 10)

gam_wf <- workflow() |>
  add_model(lm_spec) |>
  add_recipe(gam_rec)

gam_fit <- gam_wf |>
  fit(data = RWAR_dosing_lag_incdec)

gam_fit |> tidy()
gam_fit |> glance()

dat_grid_dose <- RWAR_dosing_lag_incdec |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(Polymer_dose = seq(min(RWAR_dosing_lag_incdec$Polymer_dose),
                                      max(RWAR_dosing_lag_incdec$Polymer_dose),
                                      length.out = 100))
dat_grid_temp <- RWAR_dosing_lag_incdec |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(mean_Temp = seq(min(RWAR_dosing_lag_incdec$mean_Temp), 
                         max(RWAR_dosing_lag_incdec$mean_Temp),
                                      length.out = 100))
dat_grid_pH <- RWAR_dosing_lag_incdec |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(mean_pH = seq(min(RWAR_dosing_lag_incdec$mean_pH), max(RWAR_dosing_lag_incdec$mean_pH),
                                      length.out = 100))

dat_grid_cond <- RWAR_dosing_lag_incdec |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(mean_Conductivity = seq(min(RWAR_dosing_lag_incdec$mean_Conductivity), max(RWAR_dosing_lag_incdec$mean_Conductivity),
                                      length.out = 100))
dat_grid_turb <- RWAR_dosing_lag_incdec |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(mean_Turbidity = seq(min(RWAR_dosing_lag_incdec$mean_Turbidity), max(RWAR_dosing_lag_incdec$mean_Turbidity),
                                      length.out = 100))
dat_grid_temp_lag1 <- RWAR_dosing_lag_incdec |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(Temp_lag1 = seq(min(RWAR_dosing_lag_incdec$Temp_lag1), max(RWAR_dosing_lag_incdec$Temp_lag1),
                                      length.out = 100))
dat_grid_temp_lag2 <- RWAR_dosing_lag_incdec |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(Temp_lag2 = seq(min(RWAR_dosing_lag_incdec$Temp_lag2), max(RWAR_dosing_lag_incdec$Temp_lag2),
                                      length.out = 100))
dat_grid_pH_lag1 <- RWAR_dosing_lag_incdec |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(pH_lag1 = seq(min(RWAR_dosing_lag_incdec$pH_lag1), max(RWAR_dosing_lag_incdec$pH_lag1),
                                      length.out = 100))
dat_grid_pH_lag2 <- RWAR_dosing_lag_incdec |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(pH_lag2 = seq(min(RWAR_dosing_lag_incdec$pH_lag2), max(RWAR_dosing_lag_incdec$pH_lag2),
                                      length.out = 100))
dat_grid_cond_lag1 <- RWAR_dosing_lag_incdec |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(Cond_lag1 = seq(min(RWAR_dosing_lag_incdec$Cond_lag1), max(RWAR_dosing_lag_incdec$Cond_lag1),
                                      length.out = 100))
dat_grid_cond_lag2 <- RWAR_dosing_lag_incdec |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(Cond_lag2 = seq(min(RWAR_dosing_lag_incdec$Cond_lag2), max(RWAR_dosing_lag_incdec$Cond_lag2),
                                      length.out = 100))
dat_grid_turb_lag1 <- RWAR_dosing_lag_incdec |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(Turb_lag1 = seq(min(RWAR_dosing_lag_incdec$Turb_lag1), max(RWAR_dosing_lag_incdec$Turb_lag1),
                                      length.out = 100))
dat_grid_turb_lag2 <- RWAR_dosing_lag_incdec |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(Turb_lag2 = seq(min(RWAR_dosing_lag_incdec$Turb_lag2), max(RWAR_dosing_lag_incdec$Turb_lag2),
                                      length.out = 100))

dose_pred <- augment(gam_fit, new_data = dat_grid_dose)
temp_pred <- augment(gam_fit, new_data = dat_grid_temp)
pH_pred <- augment(gam_fit, new_data = dat_grid_pH)
cond_pred <- augment(gam_fit, new_data = dat_grid_cond)
turb_pred <- augment(gam_fit, new_data = dat_grid_turb)
tempLag1_pred <- augment(gam_fit, new_data = dat_grid_temp_lag1)
tempLag2_pred <- augment(gam_fit, new_data = dat_grid_temp_lag2)
pHLag1_pred <- augment(gam_fit, new_data = dat_grid_pH_lag1)
pHLag2_pred <- augment(gam_fit, new_data = dat_grid_pH_lag2)
condLag1_pred <- augment(gam_fit, new_data = dat_grid_cond_lag1)
condLag2_pred <- augment(gam_fit, new_data = dat_grid_cond_lag2)
turbLag1_pred <- augment(gam_fit, new_data = dat_grid_turb_lag1)
turbLag2_pred <- augment(gam_fit, new_data = dat_grid_turb_lag2)

```{r}
for(pred in c("age", "year", "education")) {
  pred_dat <- dose_pred_gam[, c(pred, ".pred", ".pred_lower", ".pred_upper")] |>
    group_by_at(pred) |>
    summarise(.pred = mean(.pred), .pred_lower = mean(.pred_lower), .pred_upper = mean(.pred_upper))
  dose_dat <- RWAR_dosing_lag_incdec[, c(pred, "Ferric.chloride..mg.L")]

  if(is.numeric(dose_dat |> pull(pred))) {
    p <- dose_dat |>
      ggplot() +
      geom_point(aes_string(pred, "wage"), alpha = .3) +
      geom_line(aes_string(pred, ".pred"), data = pred_dat, colour = "darkgreen") +
      geom_line(aes_string(pred, ".pred_lower"), data = pred_dat, colour = "darkgreen", lty = 2) +
      geom_line(aes_string(pred, ".pred_upper"), data = pred_dat, colour = "darkgreen", lty = 2)
  } else {
    p <- dose_dat |>
      ggplot() +
      geom_jitter(aes_string(pred, "wage"), alpha = .3) +
      geom_point(aes_string(pred, ".pred"), data = pred_dat, colour = "darkgreen") +
      geom_errorbar(aes_string(pred, ymin = ".pred_lower", ymax = ".pred_upper"), data = pred_dat, colour = "darkgreen")
  }
  print(p)
}
```
